---
output:
  xaringan::moon_reader:
    seal: false
    includes:
      after_body: insert-logo.html
    self_contained: false
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
editor_options:
  chunk_output_type: console
---
class: center, inverse, middle

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

```{r xaringan-tile-view, echo=FALSE}
xaringanExtra::use_tile_view()
```

```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_progress_bar(color = "#808080", location = "top")
```

```{css echo=FALSE}
.pull-left {
  float: left;
  width: 44%;
}
.pull-right {
  float: right;
  width: 44%;
}
.pull-right ~ p {
  clear: both;
}


.pull-left-wide {
  float: left;
  width: 66%;
}
.pull-right-wide {
  float: right;
  width: 66%;
}
.pull-right-wide ~ p {
  clear: both;
}

.pull-left-narrow {
  float: left;
  width: 30%;
}
.pull-right-narrow {
  float: right;
  width: 30%;
}

.tiny123 {
  font-size: 0.40em;
}

.small123 {
  font-size: 0.80em;
}

.large123 {
  font-size: 2em;
}

.red {
  color: red
}

.orange {
  color: orange
}

.green {
  color: green
}
```



# Statistics
## Final Exam Information

### Christian Vedel,<br>Department of Economics<br>University of Southern Denmark

### Email: [christian-vs@sam.sdu.dk](mailto:christian-vs@sam.sdu.dk)

### Updated `r Sys.Date()`



---
class: middle
# The Final Exam

.pull-left-wide[
**Key facts:**

- **When:** June 2026. Check exam plan for time/date
- **Duration:** 2 hours
- **Format:** A few questions — written, digital exam (Digital Exam platform)
- **Location:** Exam room at the university
- **Weight:** 60% of your final grade
- **Syllabus:** The full course syllabus *(all lectures and exercises)*
- **Re-exam:** August — same format, same full syllabus
- **Formulas:** Same formulas as midterm (and previous years)
]

.pull-right-narrow[
.small123[
**Grading**
- 7-point grading scale
- Minimum **02** required to pass

**Note**
- Final grade = 40% midterm + 60% final exam
- Both parts must pass with at least **02**
]
]

---
class: middle
# Exam Rules & How to Prepare

.pull-left-wide[
**What you are allowed to use:**

- Watermarked Word or Excel file *(provided via Digital Exam)*
- Wordmat
- Maple
- Physical four-function calculator


**Practicalities:**

- Submit your watermarked file as a PDF through Digital Exam
- I will only read your PDF file — don't expect me to open Word or Excel files
- Bring your **Student ID card** / exam number
- I will not be present — invigilators are there if you need help

**Preparation:**

- Old exams are available on **ItsLearning** — use them!
- Two practice exam sets are also available — see later slide for details
]

.pull-right-narrow[
.small123[
**Questions?**

Feel free to ask me — either after class or by email:

[christian-vs@sam.sdu.dk](mailto:christian-vs@sam.sdu.dk)
]
]

---
class: middle

# Some advice

.pull-left-wide[
### For preparation
1. Set a timer and try to solve the practice exam sets *(details on a later slide)*.
2. Think of a simple template for how to answer questions.
3. Finish all exercises and participate in TA classes.

### During the exam
1. You have to communicate clearly. Write down your reasoning and steps and mark your conclusion clearly. $\leftarrow$ **Most important advice**
2. Spend 5 minutes in the beginning with slowly reading everything. Then make a plan for how to solve the questions. This will save you time in the end and make sure you don't miss anything.
3. Don't panic. It's not that important.
]

---
class: middle

# What the Exam Tests

.pull-left-wide[
- We are not testing whether you know .red[formulas] by heart.
- We are testing whether you know .red[statistics] by heart.

The key question is: **Can you navigate the central topics of the course?**

- Given a problem, can you identify what statistical concept applies?
- Can you reason clearly about uncertainty, distributions, estimation, and hypothesis testing?
- Can you manipulate the tools of statistics to solve problems and interpret results?

Formulas are provided. Understanding of theory is not provided.
]

.pull-right-narrow[
.small123[
**Central topics** are described in:
- The course description
- The lecture slides
- The textbook

These are your guide to what matters.
]
]

---
class: middle

# About This Year's Exam

.pull-left-wide[
Previous exams were **written by a different lecturer**.

**What stays the same:**
- Same topics and curriculum — nothing has changed there
- Similar question format

**What may differ:**
- You will be under a bit of time pressure. 
- There will be a few questions that are extra hard.
- I know this and adjust my assessment accordingly. You are not expected to solve every question perfectly.
- Designed to distinguish students who know their statistics *well* from those who know it *very well*

**My advice:** study the full curriculum; don't pattern-match to previous years.
]

.pull-right-narrow[
.small123[
**Bottom line:**

Understand the material. Don't over-index on specific questions from previous years.
]
]

---
class: middle

# Reading Exam Questions Carefully

.pull-left-wide[
**Read every question slowly and engage with it fully.**

Statistics is a precise subject. Questions are worded carefully, and that precision is intentional.

If something in a question seems surprising or takes you in an unexpected direction — that is worth engaging with, not dismissing.

- Do not assume a question contains an error and move on
- Engage carefully with what is actually being asked
- Your reasoning in response to a challenging question is itself part of what I am assessing

**I do not ask trick questions** — but I do ask questions that require careful thinking.
]

---
class: middle

# Practice Exam Sets

.pull-left-wide[
When writing the exam, I draft many questions and discard the ones I am not satisfied with.

I have collected **two sets of discarded questions** into practice exams — available on **ItsLearning**.

**How to use them:**

1. Set a timer and solve a full set under exam conditions *(2 hours, same aids)*
2. Try one set **early** in your reading period — to see where you stand
3. Try the other set **near the end** — to check whether you are ready

**Keep in mind:** these questions were discarded for a reason. They may be slightly off in some way. But they give you a good sense of the format and level of difficulty.
]

.pull-right-narrow[
.small123[
**Two practice sets** on ItsLearning

Use them actively — not just as reading material, but timed and written.
]
]

---
class: middle

# Use the Tools Available to You

.pull-left-wide[
**Make use of AI tools — e.g. ChatGPT — during your preparation.**

A useful workflow:

1. Upload the **lecture slides** + **course description** + **practice exams** + **exercise sets** to ChatGPT
2. Ask it to generate additional exam-style questions
3. Solve those questions yourself, then ask it to check your reasoning

This can give you a large supply of varied practice problems tailored to this course.

AI is a tool for *practice and feedback* — not a substitute for understanding. The exam tests your thinking, not your ability to prompt a model.
]

.pull-right-narrow[
.small123[
**Questions?**

[christian-vs@sam.sdu.dk](mailto:christian-vs@sam.sdu.dk)
]
]


---
class: middle

# Plan for the rest of the semester

*Lectures as planned - we are not done with your curriculum yet!*


.pull-left-wide[
...Also:
  - 2026-05-26 and/or 2026-05-29: Q&A plus recorded summary of semester lecture (both online)
  - 2026-06-02: Scheduled office hours 13:00-15:00. In my office, or if many sign up, in a larger room. Sign-up on ItsLearning.
  - 2026-06-12: Final exam (never trust a date like this in a set of slides — check the official exam plan)
]

.pull-right-narrow[
.small123[
**Questions?**

[christian-vs@sam.sdu.dk](mailto:christian-vs@sam.sdu.dk)
]
]