---
output:
  xaringan::moon_reader:
    seal: false
    includes:
      after_body: insert-logo.html
    self_contained: false
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
editor_options:
  chunk_output_type: console
---
class: center, inverse, middle

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

```{r xaringan-tile-view, echo=FALSE}
xaringanExtra::use_tile_view()
```

```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_progress_bar(color = "#808080", location = "top")
```

```{css echo=FALSE}
.pull-left {
  float: left;
  width: 44%;
}
.pull-right {
  float: right;
  width: 44%;
}
.pull-right ~ p {
  clear: both;
}


.pull-left-wide {
  float: left;
  width: 66%;
}
.pull-right-wide {
  float: right;
  width: 66%;
}
.pull-right-wide ~ p {
  clear: both;
}

.pull-left-narrow {
  float: left;
  width: 30%;
}
.pull-right-narrow {
  float: right;
  width: 30%;
}

.tiny123 {
  font-size: 0.40em;
}

.small123 {
  font-size: 0.80em;
}

.large123 {
  font-size: 2em;
}

.red {
  color: red
}

.orange {
  color: orange
}

.green {
  color: green
}
```



# Statistics
## Midterm Information

### Christian Vedel,<br>Department of Economics<br>University of Southern Denmark

### Email: [christian-vs@sam.sdu.dk](mailto:christian-vs@sam.sdu.dk)

### Updated `r Sys.Date()`



---
class: middle
# The Midterm Exam

.pull-left-wide[
**Key facts:**

- **When:** 2026-03-19. Check exam plan for time/date
- **Duration:** 1 hour
- **Format:** A few questions — written, digital exam (Digital Exam platform)
- **Location:** Exam room at the university *(like end-of-semester exam)*
- **Weight:** 40% of your final grade
- **Syllabus:** All material from lectures and exercises up to the midterm
- **Re-exam:** August — note: the re-exam covers the *full* course syllabus
- **Formulas:** Key formulas will be provided. 
]

.pull-right-narrow[
.small123[
**Grading**
- 7-point grading scale
- Minimum **02** required to pass
]
]

---
class: middle
# Exam Rules & How to Prepare

.pull-left-wide[
**What you are allowed to use:**

- Watermarked Word or Excel file *(provided via Digital Exam)*
- Wordmat
- Maple
- Physical four-function calculator


**Practicalities:**

- Submit your watermarked file as a PDF through Digital Exam
- Bring your **Student ID card** / exam number
- I will not be present — invigilators are there if you need help

**Preparation:**

- Old exams are available on **ItsLearning** — use them!
- Focus on understanding, not memorising formulas
]

.pull-right-narrow[
.small123[
**Questions?**

Feel free to ask me — either after class or by email:

[christian-vs@sam.sdu.dk](mailto:christian-vs@sam.sdu.dk)
]
]

---
class: middle

# Some advice

.pull-left-wide[
### For preparation
1. Set a timer and try to solve previous sets.
2. Practice your understanding of concepts - don't just memorise formulas.
3. Think of a simple template for how to answer questions.
4. Participate in TA classes.

### During the exam
1. You have to communicate clearly. Write down your reasoning and steps and mark your conclusion clearly. $\leftarrow$ **Most important advice**
2. Spend 5 minutes in the beginning with slowly reading everything. Then make a plan for how to solve the questions. This will save you time in the end and make sure you don't miss anything.
3. Don't panic. It's not that important. 
]
